---
title: "Module - Normalization and Dimensionality Reduction"
description: | 
  Description
execute:
  freeze: auto
  cache: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Test'
  seurat_object: './data/ovarian_qc_merged_object.RDS'
  thr_n_features: 2000
  thr_n_dimensions: 100
  n_threads: 8
  n_memory: 8
  work_directory: !expr here::here()
  auto_save: TRUE
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE, echo = FALSE}

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object

# Inputs and thresholds
thr_n_features <- params$thr_n_features
thr_n_dimensions <- params$thr_n_dimensions

# Optional parameters

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
work_directory <- params$work_directory
timestamp <- params$timestamp
auto_save <- params$auto_save

# Knitr options
knitr::opts_knit$set(
  root.dir = work_directory
  )

```

## Loading library

```{r library}
#| output: false
#| code-fold: true

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(BPCells)

```

## Folder structure

```{r project_directory}
#| output: false
#| code-fold: true

if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

# showWarnings = FALSE: data/ already exists (the BPCells store is staged there).
for(sub_directory in c('data', 'figures')) {
  dir.create(paste0(work_directory, '/', sub_directory), showWarnings = FALSE)
}

```

## Loading Seurat object

```{r loading_rds}

# The merged object is BPCells-backed. Its counts live in the on-disk store
# staged at data/bpcells_counts. Rebuild the object from that store plus the
# saved per-cell metadata so we never depend on the (possibly stale) matrix
# path baked into the RDS by the upstream task.
loaded_object <- readRDS(file = seurat_object)

disk_counts <- BPCells::open_matrix_dir(dir = file.path('data', 'bpcells_counts'))

keep_meta <- loaded_object@meta.data
keep_meta <- keep_meta[, setdiff(colnames(keep_meta),
                                 c('nCount_RNA', 'nFeature_RNA')), drop = FALSE]

seurat_object <- CreateSeuratObject(
  counts    = disk_counts,
  project   = project_name,
  meta.data = keep_meta
)

rm(loaded_object); gc()

# Setting random seed
random_seed <- 2203
set.seed(random_seed)

```

## Counts normalization

```{r normalization, message = FALSE, warning = FALSE, echo = TRUE}

seurat_object <- NormalizeData(
  object = seurat_object, 
  normalization.method = "LogNormalize",
  scale.factor = 10000)

seurat_object <- FindVariableFeatures(
  seurat_object, selection.method = "vst", nfeatures = thr_n_features)

seurat_object <- ScaleData(seurat_object)

```

## Dimensionality Reduction

```{r dimensionality_reduction}
#| warning: false

# Running PCA and printing out dimensions
seurat_object <- RunPCA(seurat_object, verbose = FALSE, npcs = thr_n_dimensions)
print(seurat_object[["pca"]], dims = 1:5, nfeatures = 5)

```

### PCA Loadings

```{r pca_loading_plot, fig.height = 14}
#| label: fig-loadings
#| fig-cap: 'PCA Loadings'
#| fig-cap-location: top
#| fig-height: 14
#| fig-width: 10

VizDimLoadings(seurat_object, dims = 1:5, reduction = "pca")

```

:::{.callout-tip}
The @fig-loadings visualizes dimensionality reduction loadings in a Seurat object. It displays the contribution of each gene to the principal components or other reduced dimensions, aiding in the interpretation of underlying biological variation.
:::

### Features Plot

```{r variable_plot}
#| label: fig-variable-plot
#| fig-cap: 'VariableFeaturePlot (TOP 10)'
#| fig-cap-location: top
#| warning: false

# Identify the 10 most highly variable genes
variable_top10 <- head(VariableFeatures(seurat_object), 10)

# plot variable features with and without labels
variable_plot <- VariableFeaturePlot(seurat_object)
variable_plot <- LabelPoints(
  plot = variable_plot, points = variable_top10, repel = TRUE)

variable_plot

```

:::{.callout-tip}
The @fig-variable-plot displays the variability of features across cells. It highlights genes with high variance, providing insights into the heterogeneity of gene expression within a dataset.
:::

### Elbow Plot

```{r elbow_plot}
#| label: fig-elbow
#| fig-cap: 'ElbowPlot'
#| fig-cap-location: top

elbow_plot <- ElbowPlot(object = seurat_object, ndims = 100)
elbow_plot

```

:::{.callout-tip}
The @fig-elbow illustrates the elbow point of the variance explained by principal components. It helps in determining the optimal number of principal components to retain for downstream analysis, such as dimensionality reduction or clustering.
:::

## Saving Seurat object

```{r object_dump}

if(auto_save) {
  
  saveRDS(seurat_object, file = paste0('data/', project_name, '_reduction_object.RDS'))
  
}

```

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

print(params)

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
